\chapter*{Acknowledgments}
%
I would like to take this opportunity to express my most sincere gratitude to all those people who have guided and supported me along my studies and in the course of my work presented in this Thesis.\medbreak

\noindent I owe a great debt of gratitude to my Thesis advisor, Emanuele R. Nocera, who really does deserve the greatest thanks for his patient guidance and willingness throughout the completion of this Thesis. I want to thank him for having carefully read the manuscript several times, providing constructive feedbacks that have been truly remarkable during this work. I have appreciated all the times he responded to my questions and queries so promptly, always exposing the topic in the clearest fashion. I will always be grateful for his overwhelming kindness and dedication to my academic growth. Thanks for having cared about my work.  
\medbreak

\noindent I am also deeply grateful to my Erasmus supervisor, Valerio Bertone, with whom I spent three wonderful months at the CEA of Paris-Saclay. I have really appreciated his commitment to share his knowledge and experience with me and for the unwavering support he provided during this project. I treasure the programming skills and the technical advice he extensively tough me. To both Emanuele and Valerio I extend my sincere thanks for having embedded me in the collaborative research environment in which I have grown not only as a (ingoing) researcher, but also as a person. Thanks for the stimulating (and funny) moments we spent together in Paris and during our weekly phone-calls.\medbreak

% University friends
\noindent I thank the theory, experiment, and astrophysics groups with whom I spent funny (though desperate) years at the Department of Physics of the University of Turin. In particular, I would like to thank Matilde Barberi Squarotti for the long study sessions she hosted at her home. Her remarkable dedication has always been an example to me.\medbreak

\noindent I am also grateful to Stefano, who made my stay in Paris even more memorable. He has always comforted me through my \quotes{existential crisis} with his unconditional patience and his immovable smile.\medbreak

\noindent I want to thank all my friends for the given support, and for having insistently made me to take some breaks from study. In particular, I would like to thank Ilaria, my sister by now, who has always loved me as if I were her brother. Thanks for having never ceased to believe in me. Additional thanks are due to Franco, Alessandro, and Marco for the long time friendship and for interesting discussions we held regarding economy, politics, and science. Thanks to Alessio, who has always been able to raise a smile from me. Were it not for your unconditional affection, these years would have been much more difficult.\medbreak

\noindent
Infine, ringrazio la mia famiglia e, in particolare, i miei genitori: Mamma, Papà e Danilo. Grazie a loro ho avuto la possibilità di frequentare l'università in completa serenità, pronti a supportarmi ogni qualvolta fosse necessario senza che gli venisse chiesto nulla. Auguro poi ad Amidia e all'ultimo arrivato Andrea di poter crescere liberi e curiosi di conoscere.

\begin{flushright}
  Amedeo Chiefa
\end{flushright}






